---
title: "02_checkm"
author: "Darius HUISMAN"
date: "2026-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

## Step 4. Check bin quality with checkm

cd to the top level of the workshop file (Workshop4_marine_genomics)
```bash
cd ~/marine-genomics/Workshop4_marine_genomics
```

Make a new directory for checkm and cd into it
```bash
mkdir checkm
cd checkm
```

Display the lengths of all contigs sorted from shortest to longest.
```bash
cat ../flye/bbd/assembly.fasta | bioawk -c fastx '{print length($seq),$name}' | sort -n
```
longest contig is contig_108

Once you have found the longest contig substitute its name instead of contig_XX in the code below.
```bash
mkdir bins_bbd
cp ../metabat/bins_bbd/*.fa bins_bbd/
samtools faidx ../flye/bbd/assembly.fasta contig_108 > bins_bbd/contig_XX.fa
```

create a slurm script for your checkm run
```bash
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=6 --mem=10gb

echo "Starting checkm in $(pwd) at $(date)"

shopt -s expand_aliases
alias checkm='apptainer run -B /pvol/:/pvol /pvol/data/sif/checkm.sif checkm'

checkm lineage_wf bins_bbd out -t 4 -x fa -f checkm_results_bbd.txt

echo "Finished checkm in $(pwd) at $(date)"
```

Submit your checkm job like this
```bash
sbatch run_checkm_bbd.sh
```


## Step 5. Select bins based on checkm results

Set directory to checkm file
```bash
cd ~/marine-genomics/Workshop4_marine_genomics/checkm
```

view the results (checkm_results_bbd.txt) in the terminal using head
```bash
head checkm_results_bbd.txt 
```
view more results
```bash
head -n 20 checkm_results_bbd.txt 
```

Extract the bin IDs for these using awk
```bash
cat checkm_results_bbd.txt | awk '$13>15'
```
Just want the bin ID 
```bash
cat checkm_results_bbd.txt | awk '$13>15{print $1}'
```
Send the output to another file for later steps
```bash
cat checkm_results_bbd.txt | awk '$13>15{print $1}' > bbd_topbins.txt
```


## Step 6. Perform taxonomic classification with gtdbtk

Create a folder for gtdbtk in the top level of the workshop file (Workshop4_marine_genomics)
```bash
cd ~/marine-genomics/Workshop4_marine_genomics
mkdir gtdbtk
cd gtdbtk
```
create a subfolder within the gtdbtk folder called genomes
```bash
mkdir genomes
```

Iterate over our list of topbins. 
(For each of the topbins we want to copy a file corresponding to that bin into our genomes folder.)
First just get the while loop working
```bash
while read bin;do echo $bin;done < ../checkm/bbd_topbins.txt
```
Turn this bin name into a path specifying the fasta file for the bin. So for every bin we need to prepend the path to its folder ../metabat/bins_bbd/ and then append the file extension .fa
```bash
while read bin;do echo "../metabat/bins_bbd/${bin}.fa";done < ../checkm/bbd_topbins.txt
```
Ready to perform the copy
```bash
while read bin;do cp "../metabat/bins_bbd/${bin}.fa" genomes;done < ../checkm/bbd_topbins.txt
```
Check that it worked by listing files within the genomes directory. (should see one file for each of the bins)
```bash
ls genomes
```

Create a blank shell script file and save it inside your gtdbtk directory, named run_gtdbtk.sh
```bash
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=2 --mem=24gb

echo "Starting gtdbtk in $(pwd) at $(date)"

shopt -s expand_aliases
alias gtdbtk='apptainer run -B /pvol:/pvol,/pvol/data/mg/gtdbtkdata/release207_v2/:/refdata /pvol/data/sif/gtdbtk.sif'

gtdbtk classify_wf -x fa --cpus 2 --genome_dir genomes --out_dir out --skip_ani_screen

echo "Finished gtdbtk in $(pwd) at $(date)"
```

Submit job
```bash
sbatch run_gtdbtk.sh
```


## Step 7. Find and annotate genes with prokka

Create a directory for our prokka work at the top of the workshop file (Workshop4_marine_genomics), and change to it.
```bash
cd ~/marine-genomics/Workshop4_marine_genomics
mkdir prokka
cd prokka
```

Find out the first bin ID in the list.
```bash
head -n 1 ../checkm/bbd_topbins.txt
```
first bin is: contig_108

Run prokka for that single bin
```bash
bin=contig_108
prokka --outdir $bin --prefix $bin  ../metabat/bins_bbd/${bin}.fa
```

List the contents of this folder.
```bash
ls contig_108
```

Run prokka on all the metagenomes.

Create a new blank text file (called it run_prokka.sh) and save it in the prokka directory.
```bash
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=2 --mem=2gb

echo "Starting prokka in $(pwd) at $(date)"

shopt -s expand_aliases
alias prokka='apptainer run -B /pvol/:/pvol /pvol/data/sif/prokka.sif prokka'

while read bin;do 
  prokka --outdir $bin --prefix $bin "../metabat/bins_bbd/${bin}.fa";
done < ../checkm/bbd_topbins.txt

echo "Finished prokka in $(pwd) at $(date)"
```

Submit job
```bash
sbatch run_prokka.sh
```
(job 605)


## Step 8. Infer metabolic pathways

Create a directory for minpath results at the top of the workshop file (Workshop4_marine_genomics), and change into it.
```bash
cd ~/marine-genomics/Workshop4_marine_genomics
mkdir minpath
cd minpath
```

Look at the .tsv file for a single bin.
```bash
head ../prokka/contig_108/contig_108.tsv
```

Transform the tsv into a format using awk to only keep a unique ID for each gene (column 1) and its EC number (column 5).
```bash
head ../prokka/contig_108/contig_108.tsv | awk -F '\t' '{print $1,$5}'
```
Filter the output to only include lines with an EC number.
```bash
head ../prokka/contig_108/contig_108.tsv | awk -F '\t' '$5{print $1,$5}'
```
Ensure that the output is tab separated and also remove the header line
```bash
head ../prokka/contig_108/contig_108.tsv | awk -F '\t' '$5{OFS="\t";print $1,$5}' | grep -v '^locus'
```
Change head to cat to include all genes
```bash
cat ../prokka/contig_108/contig_108.tsv | awk -F '\t' '$5{OFS="\t";print $1,$5}' | grep -v '^locus'
```
Send the output to a file so that it can be used as input to minpath
```bash
cat ../prokka/contig_108/contig_108.tsv | awk -F '\t' '$5{OFS="\t";print $1,$5}' | grep -v '^locus' > contig_108.ec
```
Run minpath on this file as follows
```bash
minpath -ec contig_108.ec -report contig_108.ec.report -details contig_108.ec.details
```
Take a look at the top few lines of output
```bash
head contig_108.ec.report
```

When your job is done look at the output log file
```bash
tail ../prokka/slurm-605.out
```

When prokka is done, run the following code to prepare minpath inputs for all bins.
```bash
while read bin;do 
  cat ../prokka/$bin/${bin}.tsv | awk -F '\t' '$5{printf("%s\t%s\n", $1,$5)}' | grep -v '^locus' > ${bin}.ec 
done < ../checkm/bbd_topbins.txt
```

Run minpath on every bin.
```bash
while read bin;do 
  minpath -ec ${bin}.ec -report ${bin}.ec.report -details ${bin}.ec.details
done < ../checkm/bbd_topbins.txt
```

Use the following command to search for the 'PWY-101' pathway across all the bins
```bash
grep 'PWY-101' *.ec.report
```
### Question 4.1
  How many bins contain at least one gene from this pathway?
    Five
  Which bin has all of the genes (ie a complete pathway)?
    metabat.28
  
Cross reference results with the output from gtdbtk to see the taxonomy of bins containing this pathway. 
Note that you will need to modify the commands below to use the correct bin numbers (these will be different for everyone).
```bash
grep 'metabat.28' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
```
The second pathway is a little less obviously named. It is called superpathway of tetrathionate reduction and has metacyc code PWY-5360. This pathway is important because it includes enzymes required to reduce sulfite to hydrogen sulfide. Note that sulfite acts as the terminal electron acceptor in this pathway, however, in a broader sense the term “sulfate reducing” refers to micro-organisms that can produce sulfide from a range of oxidised sulfur compounds.

Use the following command to search for the 'PWY-101' pathway across all the bins
```bash
grep 'PWY-5360' *.ec.report
```

Visualise results table for gtdbtk
```{r}
gtdbtk.results <- read_tsv("gtdbtk/out/gtdbtk.bac120.summary.tsv")
```

### Question 4.2
  How many bins contain the pathway PWY-5360?
    One (metabat.44)
  What family (in a taxonomic sense) do these bins belong to?
    Arcobacteraceae (genus Halarcobacter)










